Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-18 paper.tex
Yans updates

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
with 1,522 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width). The inset histogram shows the
distribution of these interval widths.
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
The top-ten sites by mutation count are annotated.
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
intervals and patterns of recurrent mutation along the genome in the Wide ARG
(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
of the results. To reduce the effect of artefactual recombinants, we
consider only the breakpoints associated with the 1,522 recombination nodes
that are ancestral to more than one sample. The mean length of these intervals
is 1,685 bases (median 962); the length distribution is summarised in the
inset histogram in Figure~\ref{fig:breakpoint-distribution}.

Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
Other than the boundaries happen to lie on gene breaks, not an awful lot I
think?]

% This isn't really true any more, is it?
% There is a preponderance of breakpoint intervals spanning the 3’ end of the
% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
% more closely, we find an enrichment in the number of intersecting intervals in
% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
% in the N gene following the peak in ORF8 towards the 3’ end.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/long_arg_recombination_intervals.pdf}
\caption{\label{fig:long_arg_breakpoint_distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Long ARG.
Top panel shows the intervals for 851 breakpoints associated
with 763 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width).
Other details as described in Figure~\ref{fig:breakpoint-distribution}.}
\end{figure}

>>>>

% - Distribution of breakpoint intervals
%     Study the breakpoint intervals in the Wide ARG. Because we are interested
% in the intervals, we want to be sure that these are robustly reported by the
% HMM. Hence, we focus on those that have two parents in the mirrored and
% unmirrored HMM runs, the same set of mutations, and the left and right parents
% are consistently imputed for Pango lineage status.
%     We focus on the Wide ARG here because it covers (roughly?) the same time
% period as the RIPPLES analysis.

We explore recombination signals across the SARS-CoV-2 genome by inspecting the
breakpoint intervals of the filtered recombination nodes from the wide ARG
(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}. The length
of a breakpoint interval reflects uncertainty in the estimated location of a
breakpoint. On average, the length of the breakpoint intervals is 2,895 bases
(median, 1,670 bases; range, 1 to 23,666 bases). This broad distribution of
interval length indicates that the location of a breakpoint often could not be
precisely determined given the observed genetic variation.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1769 breakpoints associated
with 1522 recombination nodes $>1$ descending samples plotted along the genome
as line segments, coloured by interval width. The inset histogram shows the
distribution of these interval widths.
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
TODO describe mutation threshold.
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
more closely, we find an enrichment in the number of intersecting intervals in
the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
4B). Markedly, there is an abrupt drop in the number of intersecting intervals
in the N gene following the peak in ORF8 towards the 3’ end.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/long_arg_recombination_intervals.pdf}
\caption{\label{fig:long_arg_breakpoint_distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Long ARG.}
\end{figure}
